---
title: "Westchester Co. NY eBird analysis"
author: "Matthew E. Aiello-Lammens"
date: "11/27/2020"
output:
  word_document: default
  html_document: default
---

Setup the analysis and read in that data

```{r}
# Call in packages
library(dplyr)
library(readr)
library(ggplot2)
library(lubridate)
library(sp)
library(raster)
library(sf)
```


```{r}
# read in data
ebd_US_NY_119_relSep_2020 <- read_delim("ebd_US-NY-119_relSep-2020/ebd_US-NY-119_relSep-2020.txt", 
                                        "\t", escape_double = FALSE, trim_ws = TRUE)
```

## Exploring the bird data

List the birds, by common name, based on most to least sitings

```{r}
bird_frequency <-
  ebd_US_NY_119_relSep_2020 %>%
  group_by(`SCIENTIFIC NAME`, `COMMON NAME`) %>%
  tally() %>% 
  arrange(desc(n))

bird_frequency
```

Let's limit our data to just the past 20 years

```{r}
# Convert obs data to a data type variable
ebd_US_NY_119_relSep_2020$`OBSERVATION DATE` = as_date(ebd_US_NY_119_relSep_2020$`OBSERVATION DATE`)

# Extract obs year
ebd_US_NY_119_relSep_2020$OBSERVATION_YEAR = year(ebd_US_NY_119_relSep_2020$`OBSERVATION DATE`)

# Filter to obs poste 2000
ebd_recent = filter(ebd_US_NY_119_relSep_2020, OBSERVATION_YEAR >= 2000)
```


Plot a histogram of observation years for the recent data

```{r}
# Look at histogram of obs years
ggplot(data = ebd_recent, aes(x = OBSERVATION_YEAR)) + 
  geom_histogram(binwidth = 1)
```


## Spatial associations

```{r}
## ---------
# Get GPS locations for field plots
field_plots = read.csv(file = "points_smpl.csv")


## Convert ebird data set into spatial points data.frame
ebd_sppnt_df = SpatialPointsDataFrame(coords = dplyr::select(ebd_recent, LONGITUDE, LATITUDE),
                                      data = ebd_recent)
crs(ebd_sppnt_df) = '+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs'

## Covert field plot locations to spatial points
field_pnts = SpatialPoints(coords = dplyr::select(field_plots, lon, lat))
crs(field_pnts) = '+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs'
```

### Read in the Westchester Co. Open Space data layer

Source: [https://gis.westchestergov.com/datasets/open-space](https://gis.westchestergov.com/datasets/open-space)

```{r}
# Get Westchester openspace layer
wc_openspace = st_read("Open_Space-shp/Open_Space.shp")

# Transform layer to WGS 84 format
wc_openspace_wgs84 <- st_transform(wc_openspace, crs = '+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs')

# Plot the layer
ggplot() +
  geom_sf(data = wc_openspace_wgs84)
```

Overlay the open space layer with the bird observation points, to associated each observation with the open space area it *may* have occurred in. 
Note that many observations are likely not from open spaces, having occurred on private property, in more developed areas, along roads, etc..

```{r}
# Match siting lat/lon with open space layer
ebd_openspace_locs = sp::over(ebd_sppnt_df, as_Spatial(wc_openspace_wgs84))

## Add open space locations with ebd data
ebd_recent <- cbind(ebd_recent, ebd_openspace_locs)
ebd_sppnt_df@data <- cbind(ebd_sppnt_df@data, ebd_openspace_locs)
```

Make a list of the number of sitings for each location identified in the open space layer

```{r}
sight_counts <-
  ebd_recent %>%
  filter(!is.na(NAME)) %>%
  group_by(NAME) %>%
  summarise(n = length(`SCIENTIFIC NAME`),
            richness = length(unique(`SCIENTIFIC NAME`))) %>% 
  arrange(desc(n))

## Plot sighting by richness to get an effort count
ggplot(data = sight_counts, aes(x = log(n), y = log(richness))) +
  geom_point()




```

```{r}
ggplot() +
  geom_sf(data = wc_openspace_wgs84) +
  geom_point(data = ebd_recent, aes(x = LONGITUDE, y = LATITUDE), alpha = 0.05)
```

Examine the spatial point distribution for the top 20 most common birds

```{r}
common_birds <- bird_frequency$`COMMON NAME`[1:20]

# Filter data set by only the 20 most common birds
ebd_recent_b <- filter(ebd_recent, `COMMON NAME` %in% common_birds)

ggplot() +
  geom_sf(data = wc_openspace_wgs84) +
  geom_point(data = ebd_recent_b, aes(x = LONGITUDE, y = LATITUDE, color = `COMMON NAME`), alpha = 0.1) +
  theme_bw()

```


Make a plot for each species

```{r}
common_bird_plots <- c()

for(x in common_birds){
  
  ebd_recent_b <- filter(ebd_recent, `COMMON NAME` == x)
  
  common_bird_plots[[x]] <-
    ggplot() +
    geom_sf(data = wc_openspace_wgs84) +
    geom_point(data = ebd_recent_b, aes(x = LONGITUDE, y = LATITUDE), alpha = 0.05)
}

common_bird_plots$`Blue Jay` +
  ggtitle("Blue Jay")
common_bird_plots$`Tufted Titmouse` +
  ggtitle("Tufted Titmouse")
```

## Birds by focal preserve

In our study, we collected vegetation and wildlife data from three Westchester preserves, and had additional data from one other preserve. 

* Teatown Lake Reservation
* Kitchawan County Park
* Ward Pound Ridge County Park
* Rockefeller State Park Preserve (additional data)

```{r}
preserves <- c("ROCKEFELLER STATE PARK PRESERVE", "TEATOWN RESERVATION", "WARD POUND RIDGE RESERVATION COUNTY PARK",
               "HUDSON PINES FARM", "KITCHAWAN PRESERVE")

ebd_recent_pres <- filter(ebd_recent, NAME %in% preserves)

# Convert Hudson Pines to Rockefeller
ebd_recent_pres$NAME[ which(ebd_recent_pres$NAME == "HUDSON PINES FARM") ] <- "ROCKEFELLER STATE PARK PRESERVE"

# Make a data.frame with preserve size and trail length information
preserve_info <- data.frame(
  NAME = c("ROCKEFELLER STATE PARK PRESERVE", "TEATOWN RESERVATION", 
           "WARD POUND RIDGE RESERVATION COUNTY PARK", "KITCHAWAN PRESERVE"),
  size_acres = c(1771, 1000, 4315, 208),
  trail_length = c(55, 15, 41.2, 5)
)

# Merge this informaiton with the ebd_recent_pres data.frame
ebd_recent_pres <- left_join(ebd_recent_pres, preserve_info, by = "NAME")

# Get observation and richness counts
preserve_sight_counts <-
  ebd_recent_pres %>%
  group_by(NAME) %>%
  summarise(size_acres = unique(size_acres),
            trail_length = unique(trail_length),
            n = length(`SCIENTIFIC NAME`),
            richness = length(unique(`SCIENTIFIC NAME`))) %>% 
  arrange(desc(n))

# Calculate miles of trail per acre of land
preserve_sight_counts$trail_per_acre = preserve_sight_counts$trail_length / preserve_sight_counts$size_acres


```

Make plots showing size / trail length relationships with counts

```{r}
ggplot(data = preserve_sight_counts, aes(x = size_acres, y = log(n))) +
  geom_point() +
  stat_smooth(method = "lm")

ggplot(data = preserve_sight_counts, aes(x = size_acres, y = richness)) +
  geom_point() +
  stat_smooth(method = "lm")

ggplot(data = preserve_sight_counts, aes(x = trail_per_acre, y = log(n))) +
  geom_point() +
  stat_smooth(method = "lm")

ggplot(data = preserve_sight_counts, aes(x = trail_per_acre, y = richness)) +
  geom_point() +
  stat_smooth(method = "lm")
```


Make some linear regression models

```{r}
summary(lm(data = preserve_sight_counts, log(richness) ~ log(size_acres)))
summary(lm(data = preserve_sight_counts, log(richness) ~ trail_per_acre))
summary(lm(data = preserve_sight_counts, log(n) ~ trail_per_acre))
```


## Scratch work

```{r}
foo = right_join(wc_openspace_wgs84, sight_counts)
ggplot() +
  geom_sf(data = foo, aes(fill = richness))
ggplot() +
  geom_sf(data = foo, aes(fill = log(n)))

```

